Dissecting the causal question underlying the association between cancer and ADRD 

### 2021-03-29 04_reports/manuscript.Rmd
monday sunshine progress

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [pdf_document]
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center")
```

@@

Pin1 is an enzime deeply involved in the pathogenesis of cancer and ADRD. One of the roles of Pin1 is regulate mitosis process and apoptosis. Over-expression of Pin1 has been found in several tumors and it's inhibition causes tumor regression. In contrast, since adult neuronal cells do not divide, Pin1 over-expression in brain tissue is used to restore tau and amyloid precursor protein to a functional shape, thus preventing the buildup of the toxic tau and beta-amyloid species associated with AD.

For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.

@@

Descriptive statistics

In a population of x, free of cancer and dementia at baseline...

We first do counts:

Time to cancer on average

We show 3 kinds, ever vs. never, time-varying cancer, and weighting by cancer dx.

We now show the risk if U, no U (Combined outcome, crude risk)

We provide additional info on causes of death for both groups, and the extreme scenarios of bounds.

Through this example we describe how much do the direct effect of cancer changes if we measure and adjust for $C$ compared to not.

@@

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [pdf_document]

---
```{r, echo =  FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align="center")
source(here::here("02_R", "03_descriptives.R"))
```

@@

Pin1 is an enzime deeply involved in the pathogenesis of cancer and ADRD. One of the roles of Pin1 is regulate mitosis process and apoptosis. Over-expression of Pin1 has been found in several tumors and it's inhibition causes tumor regression. In contrast, since adult neuronal cells do not divide, Pin1 over-expression in brain tissue is used to restore tau and amyloid precursor protein to a functional shape, thus preventing the buildup of the toxic tau and beta-amyloid species associated with AD. For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.

@@

In this section we illustrate how previous DAGS may reflect on the analysis performed to unveil the effect of Pin1 in all cause-dementia. We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands. Recruitment and initial assessments were held between 1990 and 1993, participants had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010. All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries. Date and cause of death was collected on a weekly basis via municipal population registries. Data on clinical outcomes was available until 2015.

To match the analysis to our initial question, we considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia. Out of 7983 persons who participated at baseline, `r total_n` were considered eligible. Time to cancer diagnosis, to dementia diagnosis and death status was measured for all participants. All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever ocurred first.

Participants had a mean age of `r mean_age`, and `r women_prop` were women. Further details on participants are presented in Table 1. Over follow up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`. From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`. Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry. In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.

We present the following analysis:

The difference in dementia risk treating cancer as ever vs. never (As if Pin1 and cancer dx. happened at the same time if looked retrospectively).

The difference in dementia risk considering cancer as a time-varying exposure independent of death. For example. at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.

@@

## Somewhere

Knowing the cause of death provides information about the direction of missclassification.
In patients with cancer we observed a slight % that died with dementia as a primary cause. On the other hand, among individuals free of cancer, we observed % of individuals who died with cancer as a primary cause.

In cancer you are afraid of missing dementia
in cancer free you are afraid of missing cancer dx
